---
title: "Predicting costs :: Part 2"
author: "Planning and Regional Development"
date: "11/26/2019"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
  word_document: default
---

\pagenumbering{arabic}


#1. Summary
Large projects - defined as the biggest 20 percent of 239 projects evaluated in Part 1 — and ones subjected to closed bidding processes explain a significant degree of the inaccuracy observed in agency cost estimation. Part 2 searches for predictive factors from within those two project subpopulations. It also considers the relationship between accuracy and the number of bidders, which was generally omitted from Part 1 as internal estimators do not know how many bidders will respond as they develop estimates. 

The agency's internal cost estimates predict 95 percent of variation in cost, using the second-lowest bid^[The agency's internal regulations require, in all but a handful of cases, the acceptance of the lowest bid. The Engineering Department views the second-lowest bid as a better predictive target.] as a predicting target. Yet on an absolute basis^[Mean absolute error, MAE.], however, the gap between internal estimate and second-lowest bid averages $2.7 million, or 18 percent of the average project size. Reducing this gap would provide for stronger confidence in long-range capital capacity estimates and could reduce the need for project-level change orders. 


```{r echo=FALSE, results="hide", message = FALSE, warning = FALSE}
rm(list = ls()) # clear global environment 
cat("\014") # clear the console 
options(warn=-1) # suppress annoying warnings 

options(repos=structure(c(CRAN="http://lib.stat.cmu.edu/R/CRAN/"))) # set CRAN mirror

ipak <- function(pkg){
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg)) 
        install.packages(new.pkg, dependencies = TRUE)
    sapply(pkg, require, character.only = TRUE)
}

packages = c("png","grid","broom","tidytext","stats","widyr","Matrix","ggrepel",
"MLmetrics","mltools","grid","arsenal","knitr","fastDummies","scales","hydroGOF")
ipak(packages)

library(olsrr) 
library(data.table) 
library(ggplot2) 
library(lubridate) 
library(doBy) 
library(DataCombine) 
library(dplyr) 
library(compare) 
library(StatMeasures) 
library(qwraps2) 
library(Hmisc) 
library(caret) 
library(mlbench) 
library(glmnet) 
library(tidyr) 
library(tidyverse) 
library(ggthemes) 
```

```{r, echo=FALSE, include=FALSE}
# Control font size of chunk output
def.chunk.hook = knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x = def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```


#2. Motivation

```{r echo=FALSE, results="hide", message = FALSE, warning = FALSE}
setwd("~/Dropbox/Work and research/Port Authority/construction_bids") 

bids = read.csv("./Bids_data_set.csv") 
bids$Date = as.Date(bids$Date,"%Y-%m-%d") 
```

```{r, echo=FALSE, include=TRUE}
# Balance of second bid and engineer estimate, for visualizations.
bids$bal = bids$Second.Bid - bids$Engr.Est 
# Summary stats to help visualize some obvious potential predictors. 
loc.summary = aggregate(accuracy ~ Loc, mean, data=na.omit(bids)) 
type.summary = aggregate(accuracy ~ Typeology, mean, data=na.omit(bids)) 
format.summary = aggregate(accuracy ~ Format, mean, data=na.omit(bids)) 
quintile.summary = aggregate(accuracy ~ quantile, mean, data=na.omit(bids))
quintile.summary2 = aggregate(bal ~ quantile, mean, data=na.omit(bids)) 
ld.summary = aggregate(accuracy ~ LD, mean, data=na.omit(bids)) 
```

##Project size (dollars bid)
```{r, echo=FALSE}
ggplot(bids, aes(x = as.factor(quantile), y = bal)) + 
    geom_hline(colour="dark gray", yintercept=1) +
    geom_jitter(width=0.2) +
    geom_crossbar(data=quintile.summary2, aes(ymin = bal, ymax = bal), 
                  size=1,col="red", width = .5) + 
      ylab("Second Bid minus Estimate") + 
    theme(axis.title.y=element_text(size=10)) + 
      scale_x_discrete(labels = c("$0-1.45M","$1.45-2.47M", "$2.47-$4.65M","$4.65-13.32M","$13.32M-4.8B")) + 
  scale_y_continuous(label=comma) 
```

```{r}
bids.big = subset(bids, bids$Second.Bid>=50000000) 
```

```{r, echo=FALSE, include=FALSE}
# Project size can be considered categorically. 
bids.big = bids.big %>%
  mutate(quantile = ntile(as.numeric(Second.Bid), 5))

bids.big$quantile = as.factor(bids.big$quantile) 

bids.big %>%
  select(Second.Bid, quantile) %>%
  group_by(quantile) %>%
  summarise(max = max(Second.Bid))

bids.big$quantile = ordered(bids.big$quantile, levels = 1:5)

quintile.summary.big = aggregate(bal ~ quantile, mean, data=na.omit(bids.big)) 
```

```{r, echo=FALSE, include=FALSE}
table(bids.big$Bids)
bids.big$bidders = ifelse(bids.big$Bids==2, "2", ifelse(bids.big$Bids==3, "3",
                        ifelse(bids.big$Bids==5,"5","6+"))) 
```


```{r, echo=FALSE}
ggplot(bids.big, aes(x = Second.Bid, y = bal, color=bidders)) + 
    geom_point(shape = 21, colour = "black", size = 2.5, stroke = 0.5) + 
    scale_color_brewer(palette="Greys") + 
    geom_hline(yintercept = 0, color="gray") + 
    geom_jitter(width=0.2) + 
    ylab("Second Bid minus Estimate") + 
    theme(axis.title.y=element_text(size=6), legend.position = c(.95, .5), 
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(4, 4, 4, 4)) + 
  
  
  
  theme(
  panel.background = element_rect(fill = "white",
                                colour = "white",
                                size = 0.5, linetype = "solid"),
  panel.grid.major = element_line(size = 0.25, linetype = 'solid',
                                colour = "gray95"), 
  panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                colour = "gray95")
  ) + 



    scale_x_continuous(label=comma, limits = c(0,600000000)) + 
    scale_y_continuous(label=comma, limits = c(-100000000,25000000)) + 
    ggrepel::geom_text_repel(aes(label = Proj), color = "gray50", 
                           size = 2.5, segment.color = "grey") 
    #theme_tufte()
```


```{r}
options(scipen=999)
ggplot(bids.big, aes(x = Second.Bid, y = bal, color = Bids)) + 
      scale_colour_gradient(low = "white", high = "black") +  geom_point(size=2) +
  scale_y_continuous(label=comma) + 
  scale_x_continuous(label=comma) + 
    ggrepel::geom_text_repel(aes(label = Proj), color = "grey35", size = 2.5, segment.color = "grey") 
```

